---
title: ""
format: lshtm-mrcg-html
toc: true
toc-depth: 2
execute:
  echo: true
  warning: true
  message: true
---

# Week one answers

These are the answers to the [week one homework](../homework_questions/week_one_homework.R). Every code block below has actually been run, so what you see underneath it is exactly what R prints.

There is very often more than one right way to do something in R. If your code produced the correct answer by a different route, it is not wrong.

::: callout-tip
## Read the explanations, not just the code

The code is the smaller half of this page. The point of each question was the idea behind it — what an object is, what a vector is, what R does when things do not line up. Those are what the rest of the course is built on.
:::

------------------------------------------------------------------------

## Question 1

> Calculate 3 to the power of 5. Save the answer as an object `three_power_five`

```{r}
three_power_five <- 3^5

three_power_five
```

The `^` symbol raises a number to a power, so `3^5` means 3 × 3 × 3 × 3 × 3.

The `<-` arrow is the important part here. Without it, R works out the answer, prints it to the console, and then forgets it. With it, the answer is stored under a name and appears in your Environment pane, ready to use again.

Writing the object's name on its own line, as in the second line above, tells R to print it. `print(three_power_five)` does exactly the same thing — the shorter version is just habit.

------------------------------------------------------------------------

## Question 2

> Calculate the square root of your answer to Question 1. Use the object you made to do this

```{r}
sqrt(three_power_five)
```

`sqrt()` is a **function**. The value you want it to work on goes inside the brackets.

The point of this question was to use the object rather than retyping 243. That matters more than it looks: if you later decide Question 1 should have been `3^6`, you change it in one place and everything downstream follows. Retyping numbers by hand is how analyses quietly drift out of step with themselves.

------------------------------------------------------------------------

## Question 3

> What type of object is `three_power_five`?

```{r}
str(three_power_five)
```

`str()` reports the **str**ucture of an object. Here it tells us `num` — a number.

More precisely R calls this a *double*, meaning a number that is allowed to have a decimal part, even when it does not happen to have one. R treats almost every number this way unless you specifically ask otherwise.

Object type matters far more than it might seem. It decides what R will let you do. A number can be multiplied; a character cannot, even if it looks like a number. `str()` is the quickest way to find out what you are actually holding, and it will save you a lot of confusion later.

------------------------------------------------------------------------

## Question 4

> Make a character vector which is two items long. The first item should be `"hello"` and the second `"world"`. Call it `hello_world_vector`

```{r}
hello_world_vector <- c("hello", "world")

hello_world_vector
```

`c()` combines things into a **vector**. Think of it as "combine". It is probably the function you will type most often in R.

The double quotes are what make these characters rather than object names. Without quotes, R would go looking for objects called `hello` and `world`, fail to find them, and give you an error.

Notice that R prints `[1]` at the start of the line. That is not part of your data — it is R telling you that the first item shown is item number 1. With longer vectors that wrap across several lines, those markers tell you where you are.

------------------------------------------------------------------------

## Question 5

::: callout-important
# This is not included in recent versions of the week one homework

This is from an earlier version of this homework and has been retained in case anyone is still working through that. If you have recently downloaded this you can ignore this
:::

> Print only the `"world"` item using a number

```{r}
hello_world_vector[2]
```

Square brackets pull items out of a vector by **position**. `"world"` is the second item, so `[2]` retrieves it.

R counts from 1. Many other programming languages count from 0, so if you have met one of those before, this is a common source of off-by-one confusion.

------------------------------------------------------------------------

## Question 6

::: callout-important
# This is also not included in recent versions of the week one homework

See above
:::

> Print only the `"world"` item using TRUE/FALSE

```{r}
hello_world_vector[c(FALSE, TRUE)]
```

This is the same square brackets, but fed a different kind of instruction. Instead of saying *which position*, you supply one `TRUE` or `FALSE` per item, and R keeps the ones marked `TRUE`.

This looks like a fussy way to get at "world" — and for this example it is. But it becomes one of the most useful things in R. Once you can write a condition that produces `TRUE` and `FALSE` for every row of a dataset, you can use it to select exactly the rows you want:

```{r}
my_numbers <- c(4, 8, 15, 16, 23, 42)

my_numbers > 15          # a TRUE/FALSE for every item

my_numbers[my_numbers > 15]   # keep only the TRUEs
```

That is filtering, and it is the same idea you will meet again in week 3 wearing a tidyverse costume.

::: callout-note
## Note the difference between `=` and `==`

A single `=` assigns a value. A double `==` asks a question: *is this equal to that?* and answers `TRUE` or `FALSE`.

Mixing them up is one of the most common early mistakes, and the error messages it produces are rarely helpful about it.
:::

------------------------------------------------------------------------

## Question 7

> Make a vector with the numbers 1 to 10 called `one_to_ten`. Check what type of object this is

```{r}
one_to_ten <- 1:10

one_to_ten

str(one_to_ten)
```

The colon `:` builds a run of consecutive whole numbers. `c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)` gives exactly the same result with far more typing.

`str()` reports `int` this time, for *integer* — a whole number. The colon operator produces integers, whereas typing `3^5` in Question 1 gave a `num`. In everyday use the difference rarely matters, since R converts between them as needed.

------------------------------------------------------------------------

## Question 8

> Multiply each number in this vector by two

```{r}
one_to_ten * 2
```

This is one of the things that makes R pleasant for data work. You do not have to write a loop, or repeat yourself ten times. You say "multiply this by two" and R applies it to every element.

This behaviour is called **vectorisation**, and nearly everything in R works this way. Whether your vector has 10 numbers or 10 million, the code is identical.

Note that `one_to_ten` itself has not changed. R made a new answer and printed it. To keep it you would need to assign it: `doubled <- one_to_ten * 2`.

------------------------------------------------------------------------

## Question 9

> Make a vector with the numbers 11 to 20 called `eleven_to_twenty`. Multiply the two vectors together

```{r}
eleven_to_twenty <- 11:20

one_to_ten * eleven_to_twenty
```

When two vectors of the **same length** are multiplied, R works element by element: the first item of one by the first item of the other, the second by the second, and so on.

So the first answer is 1 × 11 = 11, the second is 2 × 12 = 24, and the last is 10 × 20 = 200.

This is not how multiplying vectors works in mathematics, where you would be thinking about dot products or matrix multiplication. R's default is the simple element-by-element version, which is almost always what you want with data.

------------------------------------------------------------------------

## Question 10

> Make a vector with just the numbers 1-3 called `one_to_three`. Try to multiply this with `one_to_ten`. Can you figure out what R has done here?

```{r}
one_to_three <- 1:3

one_to_ten * one_to_three
```

**R has reused the short vector to make up the difference.** This is called **recycling**.

Ten numbers needed a partner each, but only three were supplied, so R went back to the beginning of the short vector and kept going:

| `one_to_ten`            | 1   | 2   | 3   | 4     | 5     | 6     | 7     | 8     | 9     | 10    |
|-------|-------|-------|-------|-------|-------|-------|-------|-------|-------|-------|
| `one_to_three` recycled | 1   | 2   | 3   | **1** | **2** | **3** | **1** | **2** | **3** | **1** |
| result                  | 1   | 4   | 9   | 4     | 10    | 18    | 7     | 16    | 27    | 10    |

The warning appears because 10 does not divide evenly by 3, so the recycling ran out part-way through and R finished on an incomplete pass. That untidy ending is what it is complaining about.

::: callout-important
## This is the one to remember from this week

Recycling is genuinely useful. `one_to_ten * 2` in Question 8 worked precisely because R recycled that single `2` across all ten items — you have been relying on it already without noticing.

But it is also a real hazard. Notice that R **gave you an answer anyway**. It warned, but it did not stop. If two columns of your data are accidentally different lengths, R may well recycle one of them and hand you a full set of numbers that are quietly, invisibly wrong.

And if the lengths happen to divide evenly — say 10 and 5 — you get **no warning at all**.

So: read warnings. A warning is not an error, and R will carry on regardless, but it is R telling you something did not go the way you probably intended.
:::

------------------------------------------------------------------------

## If something did not work

A few things worth checking when code will not run:

- **Spelling and capitalisation.** R thinks `one_to_ten` and `One_to_ten` are two entirely different things.
- **Unclosed brackets.** If the console shows a `+` instead of a `>`, R is still waiting for you to finish. Press <kbd>Esc</kbd> to escape and start again.
- **Missing quotes.** `"hello"` is a piece of text; `hello` is the name of an object R will go looking for.
- **Assignment.** If an object is missing from your Environment pane, check you used `<-` rather than only printing the answer.

Still stuck? Bring it to the start of the next session — the first ten minutes are set aside for exactly this.
