---
title: ""
author: Jack Goodall
format:
  lshtm-mrcg-html:
    code-fold: true
    code-summary: "Show the code"
    fig-width: 7
    fig-height: 4.2
  pdf:
    fig-width: 6.5
    fig-height: 3.9
    fig-format: pdf
    geometry:
      - top=25mm
      - bottom=25mm
      - left=22mm
      - right=22mm
toc: true
toc-depth: 3
execute:
  echo: true
  warning: false
  message: false
  fig-align: center
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(tidyr)
library(broom)
library(knitr)

# Show the code in HTML (where it is collapsed behind "Show the code"), but
# never in PDF or any other output. This has to be done here rather than in the
# YAML, because Quarto has no way to set `echo` per output format.
knitr::opts_chunk$set(echo = knitr::is_html_output())

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", colour = "#0D5257"),
      plot.subtitle = element_text(colour = "#568689"),
      panel.grid.minor = element_blank()
    )
)

teal   <- "#0D5257"
mid    <- "#568689"
pale   <- "#E2EAEB"
accent <- "#621244"
trio   <- c("#0D5257", "#568689", "#621244")

# The clean, built-in copy - identical to the course data once the `Setosi`
# typo has been corrected
iris <- datasets::iris

# --- small helpers so the tables below read consistently -------------------

fmt_p <- function(p) {
  ifelse(p < 0.001, "< 0.001", sprintf("%.3f", p))
}

# Pull the useful parts out of any t-test into one tidy row
t_row <- function(tt, label) {
  est <- if (length(tt$estimate) == 2) {
    unname(tt$estimate[1] - tt$estimate[2])
  } else {
    unname(tt$estimate)
  }
  data.frame(
    Comparison        = label,
    Estimate          = round(est, 3),
    `Lower 95%`       = round(tt$conf.int[1], 3),
    `Upper 95%`       = round(tt$conf.int[2], 3),
    `t`               = round(unname(tt$statistic), 2),
    `df`              = round(unname(tt$parameter), 1),
    `p-value`         = fmt_p(tt$p.value),
    check.names = FALSE,
    row.names = NULL
  )
}
```

# Statistics Refresher

This page is the **why** behind the week 4 session sheet. It is a companion to
that sheet, and definitely not a statistics course in itself. There are proper refresher links
on the [Resources](resources.qmd) page.

I have hidden the R code here deliberately, this is about the stats not about R, however if you are curious you can click the `show the code` buttons to see the code I used.

There are also some call out boxes below, which can be safely skipped if you are after the refresher, but which 
might tempt you in if you want some more depth. 

If you prefer a hard copy you can download a PDF on right hand side. 

::: {.callout-warning}
## A warning about stats in R

R will run almost any test you ask it to, on any data, without complaint. It has
no idea whether the test suits your question. If you are using statistical tests in your analyses
I urge you to really understand how they work, whether they are right for your data and how they should be reported.
:::

---

## Frequentist Statistics

There are two major statistical schools: frequentists and Bayesian.   Although there are some fundamental differences
much of the mathematical concepts and procedures overalap.  The test procedures discussed here are specifically ***frequentist***. 

Every test on this page works the same basic way:

1. Assume nothing interesting is going on. This is the **null hypothesis**.
2. Work out what your data would typically look like if this null hypothesis were true.
3. See how far your actual data sits from that.
4. If it sits somewhere very unlikely, doubt the assumption in step 1.

The output of step 4 is the **p-value**.

### What a p-value actually is

A p-value is the probability of getting a result at least as extreme as yours,
**assuming the null hypothesis is true**.


```{r}
#| label: fig-pvalue
#| fig-cap: "The p-value is the shaded area — the proportion of possible results at least as extreme as the one you got. Here the observed statistic is 2.1, with 20 degrees of freedom."
x    <- seq(-4, 4, length.out = 600)
df_t <- data.frame(x = x, y = dt(x, df = 20))
obs  <- 2.1

ggplot(df_t, aes(x, y)) +
  geom_area(data = subset(df_t, x >=  obs), fill = teal, alpha = 0.75) +
  geom_area(data = subset(df_t, x <= -obs), fill = teal, alpha = 0.75) +
  geom_line(linewidth = 0.9, colour = "grey25") +
  geom_vline(xintercept = c(-obs, obs), linetype = "dashed", colour = accent) +
  annotate("text", x =  3.1, y = 0.06, label = "p/2", colour = teal, fontface = "bold") +
  annotate("text", x = -3.1, y = 0.06, label = "p/2", colour = teal, fontface = "bold") +
  annotate("text", x = 0, y = 0.2,
           label = "what we'd expect\nif nothing were going on",
           colour = "grey35", size = 3.4) +
  labs(title = "A p-value is an area under a curve",
       x = "some test statistic", y = "probability density")
```

**It is not** the probability that your hypothesis is wrong. It is not the
probability the result was chance. And it is definitely not a measure of how big or
important the effect is; with a large enough sample, a trivially small
difference earns a tiny p-value, and a really important result can hide within noisy data.

There is an [excellent paper](https://link.springer.com/article/10.1007/s10654-016-0149-3) on how p-values are misinterpreted that you might enjoy. 

### Confidence intervals are usually more useful

A p-value gives you one number about *surprise*. A confidence interval gives you
a *plausible range for the actual answer*, in the units you measured. For most
reporting, this interval is more informative.

Every test below that can give you one, does.

::: {.callout-note collapse="true"}
## For the pedants: a much longer discussion of the confidence interval

"A plausible range for the answer" I've used above is the standard hand-wave description of a confidence, 
and it is loose enough to be wrong in a fundamental way.

Imagine the true parameter value you are after - say the mean
sepal length of *every* iris in the world. That number exists, it is fixed, and
you will never know it.

You run a study on 150 flowers and calculate an interval. Now imagine running
that same study again, and again, on fresh samples of 150 flowers. Each study
gives a slightly different sample, so each one produces a slightly different confidence
interval. The true value never budges; the intervals jump around it based on the nature of your sample.

> A 95% confidence intervals implies that across all those repeated
> studies, 95% of the intervals it produces would contain the true value.

Notice that this describes **the long-run
behaviour of the method**, not the particular interval sitting in your results
table.

**The three to note:**

***The interval moves, the truth doesn't.*** Your interval is calculated from your
data, so it varies from study to study. The true value is a fixed property of
the world. The statement is about how often the interval **lands on** the true
value, not about the true value falling inside your interval. Those might sound like
the same sentence, and they are not.

***It has to work whatever the truth happens to be.*** The 95% guarantee must hold
regardless of what the true value is and you don't get to know it in
order to check.

*Once you've calculated it, the probability has gone.* Your interval came out as
5.71 to 5.98. Every number there is now fixed, and so is the truth. The true
mean is either in that range or it isn't. The probability is 100% or 0%, and you have no way of
knowing which.

You will often see the line in papers: *"There is a 95% probability the
true mean lies between 5.71 and 5.98."* As I hope I have convinced you, this is false. It is a probability statement
about the true value, and in this framework the true value isn't random — it's
just unknown. That sentence describes a Bayesian **credible interval**, which is
a genuinely different thing and requires you to specify prior beliefs before you
start. (If you ask me the Bayesian approach is almost invariably more useful, but that is a story for another day).

The 95% belongs to the **procedure**, not to your interval.

::: {.callout-tip}
## An analogy

You are throwing hoops at a peg in the ground, blindfolded.

The peg does not move: that's the true value. The hoops land in different
places every throw: those are your intervals. "95% confidence" means that
across many throws, 95% of hoops ring the peg.

Taking the blindfold off after a single throw and the question "what is the
probability this hoop ringed the peg?" no longer makes sense. It did or it
didn't. The 95% was always a statement about the quality of your throwing in the longrun, not about that
particular hoop.
:::

```{r}
#| label: fig-coverage
#| fig-height: 6
#| fig-cap: "100 samples drawn from a population whose mean really is 5. Each horizontal line is one study's 95% confidence interval."
set.seed(2024)
n_sim <- 100
n_obs <- 20
mu    <- 5

sims <- do.call(rbind, lapply(seq_len(n_sim), function(i) {
  s  <- rnorm(n_obs, mean = mu, sd = 1)
  ci <- t.test(s)$conf.int
  data.frame(i = i, lo = ci[1], hi = ci[2], est = mean(s))
}))
sims$covers <- sims$lo <= mu & sims$hi >= mu

ggplot(sims, aes(y = i)) +
  geom_vline(xintercept = mu, colour = accent, linewidth = 0.9) +
  geom_linerange(aes(xmin = lo, xmax = hi, colour = covers), linewidth = 0.45) +
  geom_point(aes(x = est, colour = covers), size = 0.8) +
  scale_colour_manual(
    values = c("TRUE" = teal, "FALSE" = accent),
    labels = c("TRUE" = "contains the true mean", "FALSE" = "misses it"),
    breaks = c("TRUE", "FALSE")
  ) +
  labs(title = "The 95% belongs to the procedure, not to your interval",
       subtitle = sprintf("%d of these %d intervals miss the true value of %g",
                          sum(!sims$covers), n_sim, mu),
       x = "estimated mean", y = "simulated study", colour = NULL) +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        panel.grid.major.y = element_blank())
```

Look at any single line in that figure. Is *it* one of the 95%? There is no way
to tell from the interval itself — which is precisely why the probability
statement cannot be attached to it.

**The 95% CI is intrinsically linked to the p value** as it is exactly
the set of values that your data would *not* have rejected, had you tested each
one as a null hypothesis at p = 0.05.

Take any number inside your interval, test it as a null hypothesis, and you get
p > 0.05. Take any number outside, and you get p < 0.05. 
:::

---

## t-tests

A t-test asks whether two means are far enough apart that chance is an
unconvincing explanation.

"Far enough apart" is doing some heavy lifting, and the output of your t-test depends on three things:

- how big the difference is
- how spread out the data is
- and how much data you have

```{r}
#| label: fig-signal-noise
#| fig-cap: "The same difference in means can be convincing or unconvincing depending entirely on the spread. Both panels have true means of 5 and 6./n (Remember that the central line in a box plot shows the *median* not the *mean*."
set.seed(42)
nn <- 60
tight <- data.frame(
  g = rep(c("A", "B"), each = nn),
  v = c(rnorm(nn, 5, 0.4), rnorm(nn, 6, 0.4)),
  panel = "Tight spread — convincing"
)
wide <- data.frame(
  g = rep(c("A", "B"), each = nn),
  v = c(rnorm(nn, 5, 2.5), rnorm(nn, 6, 2.5)),
  panel = "Wide spread — unconvincing"
)

ggplot(rbind(tight, wide), aes(x = g, y = v, fill = g)) +
  geom_boxplot(alpha = 0.65, outlier.shape = NA, width = 0.55) +
  geom_jitter(width = 0.12, alpha = 0.35, size = 1) +
  facet_wrap(~panel) +
  scale_fill_manual(values = c(teal, mid)) +
  labs(title = "Same difference, different verdict",
       subtitle = "A difference is only impressive relative to the noise around it",
       x = NULL, y = "measurement") +
  theme(legend.position = "none")
```

This is why the t-statistic is essentially the *difference in means* divided by
*how variable the data is*. A large difference in noisy data can score the same
as a small difference in tidy data.

### Flavour 1: one sample

*Is this mean different from some specific number?*

Taking the iris data: is mean sepal length different from 5.5 cm?

```{r}
#| label: tbl-onesample
#| tbl-cap: "One-sample t-test of mean sepal length against 5.5 cm. The estimate here is the observed mean itself."
tt1 <- t.test(iris$Sepal.Length, mu = 5.5)
kable(t_row(tt1, "Mean sepal length vs 5.5 cm"), align = "lrrrrrr")
```

```{r}
#| label: fig-onesample
#| fig-cap: "The observed mean and its confidence interval, against the value being tested."
ggplot(iris, aes(x = Sepal.Length)) +
  geom_histogram(bins = 24, fill = pale, colour = "white") +
  annotate("rect", xmin = tt1$conf.int[1], xmax = tt1$conf.int[2],
           ymin = -Inf, ymax = Inf, fill = accent, alpha = 0.18) +
  geom_vline(xintercept = 5.5, colour = "black", linewidth = 1, linetype = "dashed") +
  geom_vline(xintercept = mean(iris$Sepal.Length), colour = accent, linewidth = 1) +
  annotate("text", x = 5.5, y = 18, label = "tested value (5.5)",
           colour = "black", hjust = 1.05, size = 3.4) +
  annotate("text", x = mean(iris$Sepal.Length)+0.05, y = 18,
           label = "observed mean\n+ 95% CI (shaded)",
           colour = accent, hjust = -0.05, size = 3.4) +
  labs(title = "Is mean sepal length different from 5.5 cm?",
       x = "Sepal length (cm)", y = "count") + 
  ylim(c(0,22))
```

The interval sits clear of 5.5, which is the same information the p-value gives
you, just expressed in centimetres rather than probability.

### Flavour 2: two independent samples

*Do these two separate groups differ?*

```{r}
#| label: tbl-twosample
#| tbl-cap: "Two-sample t-test comparing mean sepal length between two species. The estimate is the difference between the group means."
two_species <- iris |>
  filter(Species %in% c("setosa", "versicolor")) |>
  droplevels()

tt2 <- t.test(Sepal.Length ~ Species, data = two_species)
kable(t_row(tt2, "setosa − versicolor"), align = "lrrrrrr")
```

```{r}
#| label: fig-twosample
#| fig-cap: "The distributions barely overlap, which is what a very small p-value looks like."
ggplot(two_species, aes(x = Sepal.Length, fill = Species)) +
  geom_density(alpha = 0.55, colour = NA) +
  geom_vline(data = two_species |>
               group_by(Species) |>
               summarise(m = mean(Sepal.Length), .groups = "drop"),
             aes(xintercept = m, colour = Species),
             linewidth = 1, linetype = "dashed", show.legend = FALSE) +
  scale_fill_manual(values = c(teal, accent)) +
  scale_colour_manual(values = c(teal, accent)) +
  labs(title = "Sepal length: setosa vs versicolor",
       subtitle = "Dashed lines are the group means",
       x = "Sepal length (cm)", y = "density")
```

The difference is about 0.93 cm, and the interval says the true difference is
somewhere between roughly 0.81 and 1.06 cm. Note how much more that tells you
than "p < 0.001".

### Variation: Welch's or Student's?

The classic [**Student's**](https://www.physoc.org/magazine-articles/the-strange-origins-of-the-students-t-test/) t-test assumes both groups have the same variance.
**Welch's** version does not, and adjusts the degrees of freedom to compensate.
Welch's is the sensible default: it costs almost nothing when variances really
are equal, and saves you when they are not.

```{r}
#| label: tbl-welch
#| tbl-cap: "The same comparison run both ways. The estimate and interval barely move; the degrees of freedom are the giveaway."
tt2s <- t.test(Sepal.Length ~ Species, data = two_species, var.equal = TRUE)
kable(rbind(
  t_row(tt2,  "Welch (default)"),
  t_row(tt2s, "Student (equal variance assumed)")
), align = "lrrrrrr")
```


::: {.callout-note collapse="true"}
## Welch's handing of degrees of freedom

Degrees of freedom can be a bit slippery.  They essentially equates to how many independent pieces of information you have
in your sample.  This isn't *quite* your sample size, as once you have estimated your parameters, some of the information in the sample is no longer independent. 
(Think if you have 10 random numbers, and you know 9 of them AND the mean of the sample, then you can also figure out what the 10th - it isn't independent). 

A t test does just that - estimates the mean. So the degrees of freedom for a single sample t test is N - 1, 
and for a two sample t test it is N - 2.

Welch's test cacluates degrees of freedom by performing an adjustment to this 
which is a weighted compromise between the two group sizes and variance. 

```{r}
#| label: fig-welch
#| fig-cap: "Welch's correction reduces the degrees of freedom, which widens the distribution slightly and makes the test appropriately more cautious."
data.frame(
  test = c("Welch (default)", "Student (equal variance)"),
  df   = c(unname(tt2$parameter), unname(tt2s$parameter))
) |>
  ggplot(aes(x = reorder(test, df), y = df, fill = test)) +
  geom_col(width = 0.55) +
  geom_text(aes(label = round(df, 1)), hjust = -0.15, colour = "grey25") +
  coord_flip() +
  scale_fill_manual(values = c(mid, teal)) +
  ylim(0, 115) +
  labs(title = "Degrees of freedom: Welch vs Student",
       x = NULL, y = "degrees of freedom") +
  theme(legend.position = "none")
```
:::

### Variation: one-sided or two-sided?

By default a t-test asks "is there *any* difference". A one-sided test asks "is
A specifically **less** than B" — and earns a smaller p-value for it, because it
only counts extreme results in one direction.

```{r}
#| label: fig-sided
#| fig-cap: "Two-sided tests count extreme results in both tails; one-sided tests count only one. This is why a one-sided p-value is half the size."
mk <- function(side) {
  d <- data.frame(x = x, y = dt(x, 20))
  keep <- if (side == "Two-sided") subset(d, x >= obs | x <= -obs) else subset(d, x <= -obs)
  keep$panel <- side
  # The two tails are disjoint. Without a grouping variable the shading geom
  # joins them into a single polygon straight across the middle of the curve,
  # so label which tail each point belongs to.
  keep$tail  <- ifelse(keep$x < 0, "left", "right")
  d$panel <- side
  list(all = d, shade = keep)
}
a <- mk("Two-sided"); b <- mk("One-sided (less)")

# Two-sided first: it is the default, and the one-sided panel reads as
# "the same thing, but only counting one end"
lev     <- c("Two-sided", "One-sided (less)")
all_d   <- rbind(a$all,   b$all)
shade_d <- rbind(a$shade, b$shade)
all_d$panel   <- factor(all_d$panel,   levels = lev)
shade_d$panel <- factor(shade_d$panel, levels = lev)

ggplot(all_d, aes(x, y)) +
  geom_ribbon(data = shade_d, aes(ymin = 0, ymax = y, group = tail),
              fill = teal, alpha = 0.75) +
  geom_line(colour = "grey25", linewidth = 0.9) +
  facet_wrap(~panel) +
  labs(title = "Where the p-value comes from",
       x = "test statistic", y = "density")
```

Only ever use a one-sided test if you decided the direction **before** looking at
the data. Choosing afterwards is how you talk yourself into a result.

### Flavour 3: paired

*Do two measurements on the **same** subjects differ?*

The `sleep` data is the original dataset Student used in 1908: ten patients,
each of whom tried two different drugs.

```{r}
#| label: fig-paired
#| fig-cap: "Left: treating the groups as unrelated hides the pattern. Right: joining each patient's two measurements shows that nearly every individual improved."
p1 <- ggplot(sleep, aes(x = group, y = extra, fill = group)) +
  geom_boxplot(alpha = 0.6, width = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.08, size = 2, alpha = 0.7) +
  scale_fill_manual(values = c(teal, mid)) +
  labs(title = "Ignoring the pairing", x = "drug", y = "extra hours of sleep") +
  theme(legend.position = "none")

p2 <- ggplot(sleep, aes(x = group, y = extra)) +
  geom_line(aes(group = ID), colour = "grey65") +
  geom_point(aes(colour = group), size = 2.5) +
  scale_colour_manual(values = c(teal, mid)) +
  labs(title = "Respecting the pairing", x = "drug", y = NULL) +
  theme(legend.position = "none")

patchwork_available <- requireNamespace("patchwork", quietly = TRUE)
if (patchwork_available) {
  print(patchwork::wrap_plots(p1, p2, nrow = 1))
} else {
  print(p1); print(p2)
}
```

The boxplots overlap heavily and look unconvincing. But almost every individual
line slopes upwards — the *within-patient* change is remarkably consistent. It is
simply swamped by the fact that some people naturally sleep far more than others.

Pairing removes that between-patient variation entirely:

```{r}
sleep_wide <- sleep %>%
  pivot_wider(values_from = extra,
              names_from = group)

colnames(sleep_wide) <- c("ID", "drug_1", "drug_2")

#| label: tbl-paired
#| tbl-cap: "Identical numbers, analysed two ways. Pairing turns an unconvincing result into a clear one."
kable(rbind(
  t_row(t.test(sleep_wide$drug_1, sleep_wide$drug_2, data = sleep_wide, paired = TRUE),  "Paired (correct here)"),
  t_row(t.test(sleep_wide$drug_1, sleep_wide$drug_2, data = sleep_wide, paired = FALSE), "Unpaired (ignores patient)")
), align = "lrrrrrr")
```

---

## When the assumptions look shaky

t-tests assume roughly normally distributed data. The way to check is to look.

A **qq plot** sorts your data from smallest to largest and plots each value
against the value you would have expected in that position if the data really
were normal. If the two agree, the points sit on a straight line. The reference
line is drawn through the quartiles, so it is fixed by the middle of your data —
which means the tails are free to wander off it, and that is exactly what you
are looking for.

::: {.callout-important}
## Check each group separately, not the pooled data

The t-test assumes normality **within each group being compared**, not of
everything lumped together. Pooling groups that genuinely differ produces a
mixture, and a mixture is never normal — so a pooled check can condemn data that
is perfectly fine for the test you are about to run.

The middle panel below is exactly that mistake.
:::

```{r}
#| label: fig-qq
#| fig-height: 3.6
#| fig-cap: "Three qq plots. Left: a single species, behaving itself. Middle: the same measurement pooled across all three species — the S-shape is the mixture, not a problem with the data. Right: genuinely skewed data."
set.seed(7)

qq_dat <- rbind(
  data.frame(value = iris$Sepal.Length[iris$Species == "setosa"],
             which = "1. One species — normal"),
  data.frame(value = iris$Sepal.Length,
             which = "2. All three pooled — a mixture"),
  data.frame(value = rexp(150, rate = 0.4),
             which = "3. Skewed data — not normal")
)

ggplot(qq_dat, aes(sample = value)) +
  stat_qq(colour = teal, alpha = 0.55, size = 1.3) +
  stat_qq_line(colour = accent, linewidth = 0.8) +
  facet_wrap(~which, scales = "free", nrow = 1) +
  labs(title = "Reading a qq plot",
       x = "expected if normal", y = "observed value")
```

Each panel has its own signature worth learning:

**Left: on the line.** Setosa sepal length. Minor wobble in the tails is normal
in a sample of 50; you are looking for systematic curvature, not perfection.

**Middle: a gentle S.** The same measurement pooled across all three species.
Points sit above the line at the bottom and below it at the top, meaning both
tails are *shorter* than a normal distribution would give. That is the
fingerprint of a mixture: three overlapping humps make the middle broad and the
extremes scarce. Nothing is wrong with the data — the wrong thing was pooling it.

**Right: an upward curve.** Values bunched at the bottom and a long tail
stretching up. This is right skew, and it is the pattern that should make you
reach for a rank-based test or a transformation.

There are formal normality tests, but treat them with suspicion: they are
over-sensitive in large samples and under-sensitive in small ones, which is
precisely backwards from what you want. Your eyes are better.

### The rank-based twins

When normality looks doubtful, the non-parametric versions throw away the actual
values and use only their **order**.

```{r}
#| label: fig-ranks
#| fig-cap: "Ranking pulls extreme values back into line. The outlier at 40 becomes simply 'the largest', which is why rank-based tests are robust to it."
set.seed(3)
raw <- c(rnorm(11, 10, 1.5), 40)
rk  <- data.frame(
  value = c(raw, rank(raw)),
  panel = rep(c("Original values", "Ranks"), each = 12),
  y = 1
)

ggplot(rk, aes(x = value, y = y)) +
  geom_point(colour = teal, size = 3, alpha = 0.75) +
  facet_wrap(~panel, scales = "free_x", ncol = 1) +
  labs(title = "What ranking does to an outlier", x = NULL, y = NULL) +
  theme(axis.text.y = element_blank(), panel.grid.major.y = element_blank())
```

| If you would have used… | …use this instead | It compares |
|---|---|---|
| Two-sample t-test | Wilcoxon rank-sum (Mann–Whitney U) | Medians |
| Paired t-test | Wilcoxon signed-rank | Median of the differences |
| ANOVA | Kruskal–Wallis | Medians across 3+ groups |
| Pearson correlation | Spearman correlation | Ranks rather than values |

The trade-off: these are more robust, but they discard information, so when the
data really is normal they are slightly less able to detect a real difference.

---

## Comparing more than two groups

### Why not just run lots of t-tests?

Because every test you run gets its own fresh chance to throw up a false
positive. Run enough and you are nearly guaranteed one (this is **very** common in papers reporting
RNAseq data...)

```{r}
#| label: fig-familywise
#| fig-cap: "Probability of at least one false positive when running multiple independent tests at the 5% level. By 14 comparisons it is more likely than not."
fw <- data.frame(n = 1:20)
fw$p <- 1 - 0.95^fw$n

ggplot(fw, aes(n, p)) +
  geom_hline(yintercept = 0.05, linetype = "dashed", colour = "grey55") +
  geom_line(colour = teal, linewidth = 1) +
  geom_point(colour = teal, size = 2) +
  annotate("text", x = 16.5, y = 0.10, label = "the 5% you thought you had",
           colour = "grey45", size = 3.3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1)) +
  labs(title = "The multiple comparisons problem",
       subtitle = "Three groups means three pairwise tests; ten groups means forty-five",
       x = "number of tests run", y = "chance of at least one false positive")
```

### ANOVA

ANOVA asks a single question instead: *is **any** group different?* It does this
by splitting the total variation in the data into two parts.

```{r}
#| label: fig-anova
#| fig-cap: "ANOVA compares the spread of the group means around the grand mean (between-group) against the spread of points around their own group mean (within-group)."
grand <- mean(iris$Sepal.Length)
gm <- iris |>
  group_by(Species) |>
  summarise(m = mean(Sepal.Length), .groups = "drop")

set.seed(11)
iris_j <- iris
iris_j$xj <- as.numeric(iris_j$Species) + runif(nrow(iris_j), -0.18, 0.18)
iris_j <- left_join(iris_j, gm, by = "Species")

ggplot(iris_j, aes(x = xj, y = Sepal.Length)) +
  geom_segment(aes(xend = xj, yend = m), colour = "grey80", linewidth = 0.3) +
  geom_point(aes(colour = Species), alpha = 0.6, size = 1.6) +
  geom_hline(yintercept = grand, linetype = "dashed", colour = accent, linewidth = 0.9) +
  geom_segment(data = gm,
               aes(x = as.numeric(Species) - 0.3, xend = as.numeric(Species) + 0.3,
                   y = m, yend = m),
               colour = "grey20", linewidth = 1.1, inherit.aes = FALSE) +
  annotate("text", x = 0, y = grand, label = "grand mean",
           colour = accent, size = 3.3, hjust = 0, vjust = -0.6) +
  scale_x_continuous(breaks = 1:3, labels = levels(iris$Species)) +
  scale_colour_manual(values = trio) +
  labs(title = "What ANOVA is actually comparing",
       subtitle = "Grey lines = within-group variation. \nDistance of black bars from the dashed line = between-group variation",
       x = NULL, y = "Sepal length (cm)") +
  theme(legend.position = "none") 
```

If the between-group variation is large relative to the within-group variation,
the groups are probably genuinely different. That ratio is the **F statistic**.

```{r}
#| label: tbl-anova
#| tbl-cap: "ANOVA of sepal length across the three species. `Species` is the between-group row; `Residuals` is the within-group variation left over."
iris_aov <- aov(Sepal.Length ~ Species, data = iris)

tidy(iris_aov) |>
  transmute(
    Source       = ifelse(term == "Species", "Between species", "Within species (residual)"),
    df           = df,
    `Sum of sq`  = round(sumsq, 2),
    `Mean sq`    = round(meansq, 3),
    `F`          = ifelse(is.na(statistic), "", sprintf("%.1f", statistic)),
    `p-value`    = ifelse(is.na(p.value), "", fmt_p(p.value))
  ) |>
  kable(align = "lrrrrr")
```

### Tukey: but *which* groups?

A significant ANOVA tells you something differs, not what. Tukey's test makes all
the pairwise comparisons and widens the intervals to pay for having made them.

Each row is **one pair of group means subtracted from each other** not a
comparison against the grand mean.

```{r}
#| label: tbl-tukey
#| tbl-cap: "Tukey pairwise comparisons. Every interval is clear of zero, so all three species differ from each other."
tk <- as.data.frame(TukeyHSD(iris_aov)$Species)
tk$comparison <- rownames(tk)

tk |>
  transmute(
    Comparison  = gsub("-", " − ", comparison, fixed = TRUE),
    Difference  = round(diff, 3),
    `Lower 95%` = round(lwr, 3),
    `Upper 95%` = round(upr, 3),
    `p (adjusted)` = fmt_p(`p adj`)
  ) |>
  kable(align = "lrrrr")
```

```{r}
#| label: fig-tukey
#| fig-cap: "Any interval clear of the dashed zero line is a difference worth believing. All three are the same width here because the groups are equal sizes."
ggplot(tk, aes(x = diff, y = comparison)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = accent, linewidth = 0.9) +
  geom_linerange(aes(xmin = lwr, xmax = upr), colour = mid, linewidth = 1.2) +
  geom_point(colour = teal, size = 3.5) +
  geom_text(aes(label = round(diff, 2)), vjust = -1.1, colour = "grey30", size = 3.3) +
  labs(title = "Which species actually differ?",
       subtitle = "Difference in mean sepal length, with family-wise 95% intervals",
       x = "difference (cm)", y = NULL)
```

**"Family-wise"** means the 95% applies to all three intervals *jointly*. Three
separate 95% intervals would only have about a 0.95³ ≈ 86% chance of all being
right at once. Tukey widens them so the guarantee covers the whole set — which is
why a Tukey interval is wider than an ordinary t-test interval for the same pair.

---

## Categorical data

When the outcome is a count in a category rather than a number, we compare the
counts we **observed** against the counts we would **expect** if the two
variables were unrelated.

A table of counts like this is called a **contingency table**. The totals down
the edges — the **margins** — are where everything starts.

```{r}
#| label: tbl-observed
#| tbl-cap: "Observed counts: car transmission type against number of cylinders, with row and column totals."
mtcars_tab <- table(
  Transmission = ifelse(mtcars$am == 1, "Manual", "Automatic"),
  Cylinders    = mtcars$cyl
)
kable(as.data.frame.matrix(addmargins(mtcars_tab)), align = "rrrr")
```

### Where the "expected" counts come from

The margins tell us two things on their own, before we look inside the table at
all. Of the 32 cars, 13 are manual. And 11 of them have four cylinders.

Now suppose transmission and cylinder count were **completely unrelated** —
knowing one tells you nothing about the other. Then the manual cars should be
spread across the cylinder groups in the same proportions as everything else.

13 of 32 cars are manual, so 40.6% of cars overall are manual. If cylinder count
is irrelevant, then 40.6% of the eleven four-cylinder cars should be manual too:

> 11 four-cylinder cars × 40.6% manual = **4.47 cars**

Which is the same as saying:

> expected count = (row total × column total) ÷ grand total

That is all an expected count is. It is the count you would get if the **only**
things driving the table were the margins — the overall number of manuals and
the overall number of four-cylinder cars — with no relationship between them.

Doing that for every cell gives the expected table:

```{r}
#| label: tbl-expected-margins
#| tbl-cap: "Expected counts if transmission and cylinder count were unrelated. Note the margins are identical to the observed table — that is the point."
cs <- suppressWarnings(chisq.test(mtcars_tab))
kable(round(as.data.frame.matrix(addmargins(as.table(cs$expected))), 2),
      align = "rrrr")
```

Two things are worth noticing.

**The margins have not changed.** Expected counts are not a different dataset;
they are the *same* totals redistributed as evenly as those totals allow. All
the chi-squared test ever does is ask whether the actual arrangement inside the
table could plausibly have come from that even redistribution.

**Expected counts need not be whole numbers.** 4.47 cars is not a car. It is an
average over the many ways 32 cars could have fallen into the table, which is
exactly what we want to compare against.

Seen as percentages, the claim being tested becomes obvious:

```{r}
#| label: tbl-pct
#| tbl-cap: "If the two variables were unrelated, the percentage of manual cars would be the same in every column. It is not remotely the same."
pct <- prop.table(mtcars_tab, margin = 2) * 100
overall <- 100 * sum(mtcars_tab["Manual", ]) / sum(mtcars_tab)

data.frame(
  Cylinders                 = colnames(mtcars_tab),
  `Cars in group`           = as.integer(colSums(mtcars_tab)),
  `% manual observed`       = sprintf("%.1f%%", pct["Manual", ]),
  `% manual if unrelated`   = sprintf("%.1f%%", overall),
  check.names = FALSE
) |>
  kable(align = "rrrr")
```

```{r}
#| label: fig-chisq
#| fig-cap: "Observed counts against those expected if transmission and cylinder count were unrelated. Chi-squared measures the total size of those gaps."
obs <- as.data.frame(mtcars_tab)
names(obs) <- c("am", "cyl", "n")
obs$type <- "Observed"

exp_df <- as.data.frame(as.table(cs$expected))
names(exp_df) <- c("am", "cyl", "n")
exp_df$type <- "Expected if unrelated"

ggplot(rbind(obs, exp_df), aes(x = cyl, y = n, fill = type)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  facet_wrap(~am) +
  scale_fill_manual(values = c(pale, teal)) +
  labs(title = "Observed vs expected counts",
       x = "cylinders", y = "number of cars", fill = NULL) +
  theme(legend.position = "bottom")
```

### Turning the gaps into one number

For each cell, take the gap between observed and expected, square it so
over- and under-shoots don't cancel out, then divide by the expected count —
because being 3 cars out matters far more in a cell expecting 4 than in one
expecting 40.

Add those up across every cell and you have the chi-squared statistic.

```{r}
#| label: tbl-contrib
#| tbl-cap: "Each cell's contribution to the chi-squared statistic. The four-cylinder column dominates, which matches what the percentages showed."
contrib <- cs$residuals^2

kable(round(as.data.frame.matrix(addmargins(as.table(contrib))), 2),
      align = "rrrr")
```

The bottom-right corner is the chi-squared statistic itself,
`r sprintf("%.2f", unname(cs$statistic))`. This table is useful in its own right:
it tells you **which cells** are driving a significant result, which the p-value
alone never will.

### The small-counts problem

Notice that the rule of thumb is about the **expected** counts, not the observed
ones. It is the expected count that sits on the bottom of each cell's
contribution, and dividing by a small number is what makes the whole thing
unstable.

Looking back at the expected table, `r sum(cs$expected < 5)` of the
`r length(cs$expected)` cells fall below 5:

```{r}
#| label: tbl-expected
#| tbl-cap: "The expected counts again, with any cell below 5 flagged. These are the cells that make the chi-squared approximation unreliable here."
exp_flag <- round(as.data.frame.matrix(cs$expected), 2)
exp_flag[] <- lapply(exp_flag, function(col) {
  ifelse(col < 5, paste0(sprintf("%.2f", col), " ⚠"), sprintf("%.2f", col))
})
kable(exp_flag, align = "rrr")
```

Fisher's exact test is the usual resort. Rather than approximating, it works out
the probability of every possible table with these same margins and counts how
many are as extreme as the one you got. 
However many people (me included) are quite critical of it.  It assumes fixed margins, which are implausible outside of
experimental data, making the results imprecise. 

```{r}
#| label: tbl-chisq-fisher
#| tbl-cap: "The same table tested two ways. Fisher's exact test computes the probability directly instead of approximating it, so it stays valid with small counts."
ft <- fisher.test(mtcars_tab)

data.frame(
  Test      = c("Chi-squared (approximate)", "Fisher's exact"),
  Statistic = c(sprintf("%.2f", unname(cs$statistic)), "—"),
  df        = c(as.character(unname(cs$parameter)), "—"),
  `p-value` = c(fmt_p(cs$p.value), fmt_p(ft$p.value)),
  check.names = FALSE
) |>
  kable(align = "lrrr")
```


---

## Correlation

*Do two continuous variables move together?* Correlation runs from −1 to +1.

```{r}
#| label: fig-corr-scale
#| fig-cap: "What different correlation coefficients actually look like."
set.seed(19)
mk_cor <- function(r, n = 120) {
  x1 <- rnorm(n); x2 <- rnorm(n)
  y  <- r * x1 + sqrt(max(0, 1 - r^2)) * x2
  data.frame(x = x1, y = y,
             lab = paste0("r = ", sprintf("%.2f", cor(x1, y))))
}
cors <- do.call(rbind, lapply(c(-0.9, -0.4, 0, 0.4, 0.9, 0.99), mk_cor))
cors$lab <- factor(cors$lab, levels = unique(cors$lab))

ggplot(cors, aes(x, y)) +
  geom_point(colour = teal, alpha = 0.5, size = 1.3) +
  facet_wrap(~lab, nrow = 2) +
  labs(title = "A tour of correlation coefficients", x = NULL, y = NULL) +
  theme(axis.text = element_blank())
```

### Always plot it first

Anscombe's quartet is four datasets built to make exactly this point. All four
have almost identical means, variances, correlations and fitted lines.

```{r}
#| label: fig-anscombe
#| fig-cap: "Anscombe's quartet. Identical summary statistics, wildly different data. Only the first is a sensible candidate for a correlation coefficient."
ans <- data.frame(
  set = rep(paste("Dataset", 1:4), each = 11),
  x = c(anscombe$x1, anscombe$x2, anscombe$x3, anscombe$x4),
  y = c(anscombe$y1, anscombe$y2, anscombe$y3, anscombe$y4)
)

ans_lab <- ans |>
  group_by(set) |>
  summarise(r = cor(x, y), .groups = "drop") |>
  mutate(lab = paste0("r = ", sprintf("%.2f", r)))

ggplot(ans, aes(x, y)) +
  geom_smooth(method = "lm", se = FALSE, colour = accent, linewidth = 0.8) +
  geom_point(colour = teal, size = 2.2, alpha = 0.85) +
  geom_text(data = ans_lab, aes(x = 12.5, y = 4.5, label = lab),
            inherit.aes = FALSE, colour = "grey30", size = 3.4) +
  facet_wrap(~set) +
  labs(title = "Why you plot your data first",
       subtitle = "All four have the same correlation and the same fitted line")
```

Dataset 2 is a perfect curve. Dataset 3 is a perfect straight line spoiled by one
outlier. Dataset 4 is a single influential point doing all the work. A
correlation coefficient hides all three.

### Pearson or Spearman?

**Pearson** measures *straight-line* association. **Spearman** ranks the data
first, so it detects any consistently increasing relationship even when curved.

```{r}
#| label: fig-pearson-spearman
#| fig-cap: "A perfectly monotonic but curved relationship. Pearson under-reports it because it is not a straight line; Spearman sees it clearly."
set.seed(5)
cd <- data.frame(x = 1:60)
cd$y <- exp(cd$x / 12) + rnorm(60, sd = 1.2)

ggplot(cd, aes(x, y)) +
  geom_point(colour = teal, alpha = 0.75, size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = accent, linewidth = 0.8) +
  labs(title = "A relationship a straight line cannot capture",
       x = NULL, y = NULL)
```

```{r}
#| label: tbl-corr
#| tbl-cap: "The same curved data measured both ways. Spearman is close to 1 because the relationship is perfectly increasing, even though it is not a straight line."
cp <- cor.test(cd$x, cd$y, method = "pearson")
csp <- suppressWarnings(cor.test(cd$x, cd$y, method = "spearman"))

data.frame(
  Method        = c("Pearson (linear)", "Spearman (rank)"),
  Coefficient   = c(sprintf("%.3f", unname(cp$estimate)),
                    sprintf("%.3f", unname(csp$estimate))),
  `p-value`     = c(fmt_p(cp$p.value), fmt_p(csp$p.value)),
  check.names = FALSE
) |>
  kable(align = "lrr")
```

A correlation of zero means no *linear* relationship. It does not mean no
relationship — which is, again, why you plot first.

---

## Linear regression

Correlation says *these move together*. Regression says *how much y changes per
unit of x*.  It also lets you add more explanatory variables.

Many books have been written about linear regression; this is a quick outline.

The basic idea is that you have plotted one variable on the x axis, one on the y.  Your computer then tries to draw a line which minimises the *sum* of all of the *square* of the distances between the actual points and the line. 

On the image below: 

- The green points are the real data 
- The purple line is the 'best fit' line
- the thin grey lines are the differences

```{r}
#| label: fig-regression
#| fig-cap: "The fitted line is the one that makes the total of the squared vertical distances (grey) as small as possible. Those distances are the residuals."
iris_lm  <- lm(Petal.Length ~ Sepal.Length, data = iris)
iris_fit <- iris
iris_fit$pred <- predict(iris_lm)

ggplot(iris_fit, aes(Sepal.Length, Petal.Length)) +
  geom_segment(aes(xend = Sepal.Length, yend = pred),
               colour = "grey75", linewidth = 0.4) +
  geom_point(colour = teal, alpha = 0.6, size = 1.8) +
  geom_line(aes(y = pred), colour = accent, linewidth = 1.1) +
  labs(title = "Least squares",
       subtitle = sprintf("Slope: %.2f cm of petal length per cm of sepal length",
                          coef(iris_lm)[2]),
       x = "Sepal length (cm)", y = "Petal length (cm)")
```

```{r}
#| label: tbl-regression
#| tbl-cap: "Regression of petal length on sepal length."
tidy(iris_lm, conf.int = TRUE) |>
  transmute(
    Term        = ifelse(term == "(Intercept)", "Intercept", "Sepal length"),
    Estimate    = round(estimate, 3),
    `Std error` = round(std.error, 3),
    `Lower 95%` = round(conf.low, 3),
    `Upper 95%` = round(conf.high, 3),
    `p-value`   = fmt_p(p.value)
  ) |>
  kable(align = "lrrrrr")
```

Reading that table:

- **Intercept**: predicted petal length when sepal length is zero. Often
  meaningless in itself; there is no flower with zero sepal length.
- **Sepal length**: the slope. Petal length changes by this much for every extra
  centimetre of sepal length.
- **p-value**: for the null hypothesis that the slope is zero (i.e. one variable is not associated with the other)

The model explains about `r sprintf("%.0f%%", 100 * summary(iris_lm)$r.squared)`
of the variation in petal length. That figure is **R-squared**, and it is a
better guide to whether a model is *useful* than any p-value.

### Adding a grouping variable

```{r}
#| label: tbl-regression2
#| tbl-cap: "Adding species to the model. `setosa` is the baseline, so the other two rows are differences relative to it."
iris_lm2 <- lm(Petal.Length ~ Sepal.Length + Species, data = iris)

tidy(iris_lm2, conf.int = TRUE) |>
  transmute(
    Term = recode(term,
                  "(Intercept)"       = "Intercept (setosa)",
                  "Sepal.Length"      = "Sepal length",
                  "Speciesversicolor" = "versicolor vs setosa",
                  "Speciesvirginica"  = "virginica vs setosa"),
    Estimate    = round(estimate, 3),
    `Lower 95%` = round(conf.low, 3),
    `Upper 95%` = round(conf.high, 3),
    `p-value`   = fmt_p(p.value)
  ) |>
  kable(align = "lrrrr")
```

There is no row for setosa because setosa **is** the reference point. Every
categorical variable in a model works this way: one level becomes the baseline,
and the others are reported as differences from it.

```{r}
#| label: fig-regression-species
#| fig-cap: "Left: one line for everyone. Right: a separate line per species — each much shallower than the overall line."
q1 <- ggplot(iris, aes(Sepal.Length, Petal.Length)) +
  geom_point(alpha = 0.55, colour = teal, size = 1.7) +
  geom_smooth(method = "lm", se = FALSE, colour = accent) +
  labs(title = "Ignoring species", x = "Sepal length (cm)", y = "Petal length (cm)")

q2 <- ggplot(iris, aes(Sepal.Length, Petal.Length, colour = Species)) +
  geom_point(alpha = 0.7, size = 1.7) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_colour_manual(values = trio) +
  labs(title = "Splitting by species", x = "Sepal length (cm)", y = NULL) +
  theme(legend.position = "bottom")

if (patchwork_available) {
  print(patchwork::wrap_plots(q1, q2, nrow = 1))
} else {
  print(q1); print(q2)
}
```

This is worth staring at. The overall line is much steeper than any of the three
individual lines, because it is partly measuring the fact that the species differ
from each other rather than the relationship within any of them. When a
relationship changes (or reverses) once you account for a group, that is
**Simpson's paradox**, and it is one of the most common ways to be badly wrong
with a perfectly valid p-value.

---

## A closing warning

Every figure and table on this page was produced without a murmur of complaint.
The same software would equally happily have run a paired test on unpaired data,
a t test on highly skewed data, or a correlation on Anscombe's fourth
dataset.

Choosing the test is your job. The coding was never the hard part.
